---
title: "Lab 1, Short Question"
geometry: margin=1in
output:
  pdf_document:
    toc: yes
    number_sections: yes
fontsize: 11pt
---

```{r load packages for homework 2, message=FALSE, warning=FALSE}
library(tidyverse)
library(stargazer)
library(finalfit)
library(nnet)
library(car)
library(ordinal)
theme_set(theme_bw()) # set the theme (theme_set is built inside ggplot2)
```

# Political ideology (30 points)

These questions are based on Question 14 of Chapter 3 of the textbook "Analysis of Categorical Data with R" by Bilder and Loughin.

> An example from Section 4.2.5 examines data from the 1991 U.S. General Social Survey that cross-classifies people according to

> - Political ideology: Very liberal (VL), Slightly liberal (SL),Moderate (M), Slightly conservative (SC), and Very conservative (VC)
> - Political party: Democrat (D) or Republican (R)
> - Gender: Female (F) or Male (M).

> Consider political ideology to be a response variable, and political party and gender to be explanatory variables. The data are available in the file pol_ideol_data.csv.

## Recode Data (2 points)

Use the factor() function with the ideology variable to ensure that R places the levels of the ideology variable in the correct order.

```{r read in data}
pol_ideol <- read.csv("pol_ideol_data.csv",
                         header=T,na.strings=c("","NA")) 
```


```{r recode data}
pol_ideol$ideol_levels <- factor(pol_ideol$ideol, levels = c("VL", "SL", "M", "SC", "VC"))
pol_ideol$gender <- as.factor(pol_ideol$gender)
pol_ideol$party <- as.factor(pol_ideol$party)

# create new column to identify combos of party and gender
pol_ideol$segment <- ifelse(pol_ideol$gender == 'F' & pol_ideol$party == 'D', "F/D", NA)
pol_ideol$segment <- ifelse(pol_ideol$gender == 'F' & pol_ideol$party == 'R', "F/R", pol_ideol$segment)
pol_ideol$segment <- ifelse(pol_ideol$gender == 'M' & pol_ideol$party == 'D', "M/D", pol_ideol$segment)
pol_ideol$segment <- ifelse(pol_ideol$gender == 'M' & pol_ideol$party == 'R', "M/R", pol_ideol$segment)
# head(pol_ideol)
```


## Test for Independence (5 points)

Analyze the relationships between political ideology and political party and gender using basic visualizations. Afterward, generate a contingency table and assess the independence of political ideology from political party and gender.

```{r Create a bar plot using the count variable, include = FALSE}
p <- ggplot(data=pol_ideol, aes(x=ideol_levels, y=count))+
  geom_bar(stat='identity') +
  ggtitle("Political Ideology Distribution") + 
  xlab("Ideology") + ylab("Frequency")
p
```

```{r include = FALSE}
p <- ggplot(data=pol_ideol, aes(x=party, y=count))+
  geom_bar(stat='identity') +
  ggtitle("Party Distribution") + 
  xlab("Ideology") + ylab("Frequency")
p
```

```{r include = FALSE}
p <- ggplot(data=pol_ideol, aes(x=gender, y=count))+
  geom_bar(stat='identity') +
  ggtitle("Gender Distribution") + 
  xlab("Ideology") + ylab("Frequency")
p
```

```{r gender-party divide for each ideology }
ggplot(pol_ideol, aes(fill=party, y=count, x=ideol_levels)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Party and Ideologies") +
    xlab("Ideologies")

ggplot(pol_ideol, aes(fill=gender, y=count, x=ideol_levels)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Gender and Ideologies") +
    xlab("Ideologies")
```

```{r}
#Plots of party and idology with gender identified in each bar plot
# variable names updated to match annie's code
p1<- pol_ideol %>%
  ggplot(aes(x=ideol_levels, fill = gender)) + geom_bar() +  labs(title = "Bar Plot of Ideology Categories")
p1
p2 <- pol_ideol %>% 
  ggplot(aes(x=party_levels, fill = gender)) + geom_bar() +  labs(title = "Bar Plot of Party Categories")
p2
```

```{r}
# barplot of relative ideology counts for each gender/party category
p1<- pol_ideol %>%
  ggplot(aes(x=segment, y= count, fill = ideol_levels)) + geom_bar(stat = "identity") +  labs(title = "Bar Plot of Ideology Categories by gender and party")
p1
```

```{r Create a contingency table and test for the independence}
# independence of political ideology and gender
tab1 <- xtabs(count ~ ideol_levels + gender, data=pol_ideol)
tab1 # contingency table ideol and gender

test1<- chisq.test(tab1, correct=FALSE) ## chi-square test
# test1$stdres
test1

# independence of political ideology and party
tab2 <- xtabs(count ~ ideol_levels + party, data=pol_ideol)
tab2 # contingency table idol an party

test1<- chisq.test(tab2, correct=FALSE) # chi-square test
# test1$stdres
test1

```

The p-values for gender and party are both small, so the null hypothesis is that party and gender are independent from ideology is rejected, and we can conclude that ideology is dependent on these two variables.

## Regression analysis  (5 points)

Estimate a multinomial regression model and ordinal (proportional odds) regression model that both include party, gender, and their interaction. Perform Likelihood Ratio Tests (LRTs) to test the importance of each explanatory variable.

Also, test whether the proportional odds assumption in the ordinal model is satisfied. Based on this test and other results, which model do you think is more valid?
  
```{r multinomial and ordinal regressions}
# multinomial model
mod.nomial <- multinom(ideol_levels ~ party + gender + party:gender, data=pol_ideol, weights = count)
summary(mod.nomial)
# ordinal model
mod.ord <- clm(ideol_levels ~ party + gender, data=pol_ideol, weights = count)
summary(mod.ord)
```


```{r LRTs}
# anova test on multinomial model
Anova(mod.nomial, test = "LR")
# anova test on ordinal model
Anova(mod.ord, test = "Chisq")
```

Using Anova to test each of the explanatory variables, the p-value is small for party, so we can conclude that is has an effect on ideology. The p-values were large for gender and the party-gender interaction effect, so for these variables we are unable to reject the null hypothesis.

For the ordinal model, the p-value for party, gender and its interaction was small, so these variables do have an affect on ideology under this model.

```{r proportional odds assumption}
# nominal test for proportional odds assumption 
nominal_test(mod.ord)
```

The p-values for all of the variables are large, we cannot reject the null hypothesis for proportional odds, so the ordinal model is not valid and we should use the multinomial model.  

## Estimated probabilities  (5 points)

Compute the estimated probabilities for each ideology level given all possible combinations of the party and gender levels.

```{r estimated probabilities for each ideology level}
# predict probability of all ideologies for female & democrat
data.f.d <- data.frame(party = 'D', gender = 'F')
pred.f.d <- predict(mod.nomial, newdata = data.f.d, type="probs", se=TRUE)

# predict probability of all ideologies for female & republican
data.f.r <- data.frame(party = 'R', gender = 'F')
pred.f.r <- predict(mod.nomial, newdata = data.f.r, type="probs", se=TRUE)

# predict probability of all ideologies for male & democrat
data.m.d <- data.frame(party = 'D', gender = 'M')
pred.m.d <- predict(mod.nomial, newdata = data.m.d, type="probs", se=TRUE)

# predict probability of all ideologies for male & republican
data.m.r <- data.frame(party = 'R', gender = 'M')
pred.m.r <- predict(mod.nomial, newdata = data.m.r, type="probs", se=TRUE)

# predicted probabilities together into table
tab_dat <- rbind(pred.f.d, pred.f.r, pred.m.d, pred.m.r)
colnames(tab_dat) <- c("VL", "SL", "M", "SC", "VC")
rownames(tab_dat) <- c("Female, Democrat", "Female, Republican", "Male, Democrat", "Male, Republican")
tab <- as.table(tab_dat)
tab
```

## Contingency table of estimated counts (5 points)

Construct a contingency table with estimated counts from the model. These estimated counts are found by taking the estimated probability for each ideology level multiplied by their corresponding number of observations for a party and gender combination.

For example, there are 264 observations for gender = "F" and party = "D". Because the multinomial regression model results in $\hat{\pi}_{VL} = 0.1667$, this model’s estimated count is $0.1667 \times 264 = 44$.

- Are the estimated counts the same as the observed? Conduct a goodness of fit test for this and explain the results.

```{r a contingency table with estimated counts}
# get observed total counts for 4 cross-categories
fd.sum <- sum(pol_ideol[which(pol_ideol$gender == "F" & pol_ideol$party == 'D'), ]$count)
fr.sum <- sum(pol_ideol[which(pol_ideol$gender == "F" & pol_ideol$party == 'R'), ]$count)
md.sum <- sum(pol_ideol[which(pol_ideol$gender == "M" & pol_ideol$party == 'D'), ]$count)
mr.sum <- sum(pol_ideol[which(pol_ideol$gender == "M" & pol_ideol$party == 'R'), ]$count)

# VL contingency
vl.prob <- unname(tab[, 1])
vl.dat <- matrix(round(c(vl.prob[1]*fd.sum, vl.prob[2]*fr.sum, vl.prob[3]*md.sum, vl.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(vl.dat) <- c("F", "M")
rownames(vl.dat) <- c("D", "R")

# SL contingency
sl.prob <- unname(tab[, 2])
sl.dat <- matrix(round(c(sl.prob[1]*fd.sum, sl.prob[2]*fr.sum, sl.prob[3]*md.sum, sl.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(sl.dat) <- c("F", "M")
rownames(sl.dat) <- c("D", "R")

# M contingency
m.prob <- unname(tab[, 3])
m.dat <- matrix(round(c(m.prob[1]*fd.sum, m.prob[2]*fr.sum, m.prob[3]*md.sum, m.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(m.dat) <- c("F", "M")
rownames(m.dat) <- c("D", "R")

# SC contingency
sc.prob <- unname(tab[, 4])
sc.dat <- matrix(round(c(sc.prob[1]*fd.sum, sc.prob[2]*fr.sum, sc.prob[3]*md.sum, sc.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(sc.dat) <- c("F", "M")
rownames(sc.dat) <- c("D", "R")

# VC contingency
vc.prob <- unname(tab[, 5])
vc.dat <- matrix(round(c(vc.prob[1]*fd.sum, vc.prob[2]*fr.sum, vc.prob[3]*md.sum, vc.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(vc.dat) <- c("F", "M")
rownames(vc.dat) <- c("D", "R")

# print all contingency tables
print("VL contingency table")
vl.dat
print("SL contingency table")
sl.dat
print("M contingency table")
m.dat
print("SC contingency table")
sc.dat
print("VC contingency table")
vc.dat

```

```{r include=FALSE}
# checking calculations on contingency table
# fd.sum
# fr.sum
# md.sum
# mr.sum
# 
# pred.f.d*fd.sum
# pred.m.d*md.sum
```

```{r goodness of fit test for observed vs. expected}
# GoF test VL observed with expected
vl.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'VL'), ]$count, nrow=2, ncol=2)
chisq.test(vl.dat, vl.actual)

# GoF test SL observed with expected
sl.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'SL'), ]$count, nrow=2, ncol=2)
chisq.test(sl.dat, sl.actual)

# GoF test M observed with expected
m.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'M'), ]$count, nrow=2, ncol=2)
chisq.test(m.dat, m.actual)

# GoF test SC observed with expected
sc.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'SC'), ]$count, nrow=2, ncol=2)
chisq.test(sc.dat, sc.actual)

# GoF test VC observed with expected
vc.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'VC'), ]$count, nrow=2, ncol=2)
chisq.test(vc.dat, vc.actual)
```

The p-values are large which means that there is not enough evidence to reject the null hypothesis and conclude that the distribution of the expected and actual counts are different from each other. The estimated counts are the same as the observed counts

## Odds ratios and confidence intervals  (8 points)

To better understand relationships between the explanatory variables and the response, compute odds ratios and their confidence intervals from the estimated models and interpret them.

```{r Odds ratios and confidence intervals}
beta.hat2 <- coefficients(mod.nomial)[1, 2:4]
beta.hat3 <- coefficients(mod.nomial)[2, 2:4]
beta.hat4 <- coefficients(mod.nomial)[3, 2:4]
beta.hat5 <- coefficients(mod.nomial)[4, 2:4]

# Sl vs VL OR
round(exp(beta.hat2), 2)

# Sl vs VL OR
round(exp(beta.hat3), 2)

# Sl vs VL OR
round(exp(beta.hat4), 2)

# Sl vs VL OR
round(exp(beta.hat5), 2)


# beta confidence interval
conf.beta <- confint(object = mod.nomial, level=0.95)

## construct CI for OR

exp(conf.beta)

```

