---
title: "Lab 1, Short Question"
geometry: margin=1in
output:
  pdf_document:
    toc: yes
    number_sections: yes
fontsize: 11pt
---

```{r load packages for homework 2, message=FALSE, warning=FALSE}
library(tidyverse)
library(stargazer)
theme_set(theme_bw()) # set the theme (theme_set is built inside ggplot2)
```

# Political ideology (30 points)

These questions are based on Question 14 of Chapter 3 of the textbook "Analysis of Categorical Data with R" by Bilder and Loughin.

> An example from Section 4.2.5 examines data from the 1991 U.S. General Social Survey that cross-classifies people according to

> - Political ideology: Very liberal (VL), Slightly liberal (SL),Moderate (M), Slightly conservative (SC), and Very conservative (VC)
> - Political party: Democrat (D) or Republican (R)
> - Gender: Female (F) or Male (M).

> Consider political ideology to be a response variable, and political party and gender to be explanatory variables. The data are available in the file pol_ideol_data.csv.

## Recode Data (2 points)

Use the factor() function with the ideology variable to ensure that R places the levels of the ideology variable in the correct order.

```{r recode data}
voters <- read.csv("pol_ideol_data.csv")

voters$ideol.order <- factor(voters$ideol, levels = c("VL", "SL", "M", "SC", "VC"))
voters
levels(voters$ideol.order)

```

## Test for Independence (5 points)

Analyze the relationships between political ideology and political party and gender using basic visualizations. Afterward, generate a contingency table and assess the independence of political ideology from political party and gender.




```{r Create a bar plot using the count variable}


ggplot(voters, aes(fill=party, y=count, x=ideol.order)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Party and Ideologies") +
    xlab("Ideologies")

ggplot(voters, aes(fill=gender, y=count, x=ideol.order)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Gender and Ideologies") +
    xlab("Ideologies")

```





```{r Create a contingency table and test for the independence}
vp.table = xtabs(formula = count ~ ideol.order + party, data = voters)
vp.table
ind.test1 <- chisq.test(x = vp.table, correct = FALSE)
ind.test1


vg.table = xtabs(formula = count ~ ideol.order + gender, data = voters)
vg.table
ind.test2 <- chisq.test(x = vg.table, correct = FALSE)
ind.test2

```

## Regression analysis  (5 points)

Estimate a multinomial regression model and ordinal (proportional odds) regression model that both include party, gender, and their interaction. Perform Likelihood Ratio Tests (LRTs) to test the importance of each explanatory variable.

Also, test whether the proportional odds assumption in the ordinal model is satisfied. Based on this test and other results, which model do you think is more valid?
  
```{r multinomial and ordinal regressions}
multi.mod <- multinom(formula = ideol.order ~ gender + party + gender:party, data = voters, weights = count)
pol.mod <-  polr(formula = ideol.order ~ gender + party + gender:party, data = voters, weights = count)
summary(multi.mod)
summary(pol.mod)
```

```{r LRTs}
Anova(multi.mod, test = "LR")
Anova(pol.mod, test = "LR")
```

```{r proportional odds assumption}
library(ordinal)
mod.ordtest <- clm(ideol.order ~ party + gender, data =voters, weights = count)

nominal_test(mod.ordtest)
```

## Estimated probabilities  (5 points)

Compute the estimated probabilities for each ideology level given all possible combinations of the party and gender levels.

```{r estimated probabilities for each ideology level}
v.table <- xtabs(formula = count ~ ideol.order + gender + party, data = voters)
v.table/rowSums(v.table)
```

## Contingency table of estimated counts (5 points)

Construct a contingency table with estimated counts from the model. These estimated counts are found by taking the estimated probability for each ideology level multiplied by their corresponding number of observations for a party and gender combination.

For example, there are 264 observations for gender = "F" and party = "D". Because the multinomial regression model results in $\hat{\pi}_{VL} = 0.1667$, this model’s estimated count is $0.1667 \times 264 = 44$.

- Are the estimated counts the same as the observed? Conduct a goodness of fit test for this and explain the results.

```{r}
cont.table <- xtabs(formula = count ~ party + ideol.order + gender, data = voters)
cont.table
flattened <- ftable(x = cont.table, row.vars = c("gender", "party"), col.vars = "ideol.order")
flatdata <- as.data.frame(flattened)
pihat <- predict(object = multi.mod, newdata = flatdata[1:4,], type = "probs")
estprob <- data.frame(gender = flatdata[1:4,], party = flatdata[1:4, 2], round(pihat, 4))
estcounts <- estprob[, 6:10] * rowSums(flattened[, 0:5])
est_counts_labeled <- data.frame(gender = flatdata[1:4,1], party = flatdata[1:4, 2], round(estcounts, 4))
est_counts_labeled
```






```{r a contingency table with estimated counts}
pihat
estprob
#v.table
#coefficients(multi.mod)
#pi.hat.ord <- predict(object = multi.mod, type = "probs")
#pi.hat.ord
```

```{r goodness of fit test for observed vs. expected}
#chisq test
chisq.test(x = estcounts, correct = FALSE)
```

## Odds ratios and confidence intervals  (8 points)

To better understand relationships between the explanatory variables and the response, compute odds ratios and their confidence intervals from the estimated models and interpret them.

```{r Odds ratios and confidence intervals}
exp(pol.mod$coefficients)
cip <- confint(object = pol.mod, level = 0.95)
exp(cip)


exp(coef(multi.mod))
cim <- confint(object = multi.mod, level = 0.95)
exp(cim)
```
