dd---
title: 'Statistical Methods for Discrete Response, Time Series, and Panel Data (W271): Lab 2'
geometry: margin=1in
output:
  github_document: default
---

# The Keeling Curve

In the 1950s, the geochemist Charles David Keeling observed a seasonal pattern in the amount of carbon dioxide present in air samples collected over the course of several years. He was able to attribute this pattern to the variation in global rates of photosynthesis throughout the year, caused by the difference in land area and vegetation cover between the Earth's northern and southern hemispheres. 

In 1958 Keeling began continuous monitoring of atmospheric carbon dioxide concentrations from the Mauna Loa Observatory in Hawaii and soon observed a trend increase carbon dioxide levels in addition to the seasonal cycle. He was able to attribute this trend increase to growth in global rates of fossil fuel combustion. This trend has continued to the present, and is known as the "Keeling Curve."

```{r load packages, echo = FALSE, message = FALSE}
library(tidyverse)
library(tsibble)
library(latex2exp)

library(feasts)
library(forecast)
library(fable)
theme_set(theme_minimal())
knitr::opts_chunk$set(dpi=1000)
```


```{r plot the keeling curve, echo = FALSE}
tsibble::as_tsibble(co2) %>%
  ggplot() + 
  aes(x=index, y=value) + 
  geom_line(color = 'steelblue') +
  labs(
    title = TeX(r'(Monthly Mean $CO_2$)'),
    subtitle = 'The "Keeling Curve"',
    x = 'Month and Year',
    y = TeX(r'($CO_2$ parts per million)')
  )
```
\newpage

# Your Assignment 

Your goal in this assignment is to produce a comprehensive analysis of the Mona Loa CO2 data that you will be read by an interested, supervising data scientist. Rather than this being a final report, you might think of this as being a contribution to your laboratory. You and your group have been initially charged with the task of investigating the trends of global CO2, and told that if you find "anything interesting" that the team may invest more resources into assessing the question. 

Because this is the scenario that you are responding to: 

1. Your writing needs to be clear, well-reasoned, and concise. Your peers will be reading this, and you have a reputation to maintain.
2. Decisions that you make for your analysis need also be clear and well-reasoned. While the main narrative of your deliverable might only present the modeling choices that you determine are the most appropriate, there might exist supporting materials that examine what the consequences of other choices would be. As a concrete example, if you determine that a series is an AR(1) process your main analysis might provide the results of the critical test that led you to that determination and the results of the rest of the analysis under AR(1) modeling choices. However, in an appendix or separate document that is linked in your main report, you might show what a MA model would have meant for your results instead.
3. Your code and repository are a part of the deliverable. If you were to make a clear argument that this is a question worth pursuing, but then when the team turned to continue the work they found a repository that was a jumble of coding idioms, version-ed or outdated files, and skeletons it would be a disappointment.

# Report from the Point of View of 1997 

For the first part of this task, suspend reality for a short period of time and conduct your analysis from the point of view of a data scientist doing their work in the early months of 1998. Do this by using data that is included in _every_ R implementation, the `co2` dataset. This dataset is lazily loaded with every R instance, and is stored in an object called `co2`. 

## (3 points) Task 0a: Introduction 

Introduce the question to your audience. Suppose that they _could_ be interested in the question, but they don't have a deep background in the area. What is the question that you are addressing, why is it worth addressing, and what are you going to find at the completion of your analysis. Here are a few resource that you might use to start this motivation. 

- [Wikipedia](https://en.wikipedia.org/wiki/Keeling_Curve)
- [First Publication](./background/keeling_tellus_1960.pdf)
- [Autobiography of Keeling](./background/keeling_annual_review.pdf)

## (3 points) Task 1a: CO2 data
Conduct a comprehensive Exploratory Data Analysis on the `co2` series. This should include (without being limited to) a [description of how, where and why ](https://gml.noaa.gov/ccgg/about/co2_measurements.html) the data is generated, a thorough investigation of the trend, seasonal and irregular elements. Trends both in levels and growth rates should be discussed (consider expressing longer-run growth rates as annualized averages).

What you report in the deliverable should not be your own process of discovery, but rather a guided discussion that you have constructed so that your audience can come to an understanding as succinctly and successfully as possible. This means that figures should be thoughtfully constructed and what you learn from them should be discussed in text; to the extent that there is _any_ raw output from your analysis, you should intend for people to read and interpret it, and you should write your own interpretation as well. 


```{r load data into tsibble}
df_co2 <- as_tsibble(co2)
```


```{r create season column}
df_co2$season <- ifelse(month(df_co2$index)== 3 | month(df_co2$index)== 4 | month(df_co2$index)== 5, "spring", NA)
df_co2$season <- ifelse(month(df_co2$index)== 6 | month(df_co2$index)== 7 | month(df_co2$index)== 8, "summer", df_co2$season)
df_co2$season <- ifelse(month(df_co2$index)== 9 | month(df_co2$index)== 10 | month(df_co2$index)== 11, "fall", df_co2$season)
df_co2$season <- ifelse(month(df_co2$index)== 12 | month(df_co2$index)== 1 | month(df_co2$index)== 2, "winter", df_co2$season)

df_co2$season <- as.factor(df_co2$season)
head(df_co2)
```

## (3 points) Task 2a: Linear time trend model

Fit a linear time trend model to the `co2` series, and examine the characteristics of the residuals. Compare this to a quadratic time trend model. Discuss whether a logarithmic transformation of the data would be appropriate. Fit a polynomial time trend model that incorporates seasonal dummy variables, and use this model to generate forecasts to the year 2020. 

```{r}
#### 2A NORA WORK FROM 3/5

# Fit a linear time trend model
linear_model <- tslm(co2 ~ trend)

# Examine characteristics of residuals
residuals_linear <- residuals(linear_model)
plot(residuals_linear, type = "l", main = "Residuals of Linear Model")

# Fit a quadratic time trend model
quadratic_model <- df_co2 %>% model(trend = TSLM(value ~ trend() + I(trend()^2)))

# Examine characteristics of residuals
residuals_quadratic <- residuals(quadratic_model)
#plot(residuals_quadratic, type = "l", main = "Residuals of Quadratic Model")

# Test for logarithmic transformation
log_model <- lm(log(co2) ~ time(co2))
summary(log_model)


# Fit a polynomial time trend model with seasonal dummies

    
poly_terms <- poly(co2, degree = 12)

# adding seasonal dummy variable to code update 3/8
# try different degrees??
polynomial_model <- df_co2 %>% model(trend = TSLM(value ~ trend() + I(trend()^2) + I(trend()^3) + season()))

residuals_poly <- residuals(polynomial_model)$.resid



```



## (3 points) Task 3a: ARIMA times series model 

Following all appropriate steps, choose an ARIMA model to fit to the series. Discuss the characteristics of your model and how you selected between alternative ARIMA specifications. Use your model (or models) to generate forecasts to the year 2022. 

```{r}
# 4 core plots
plot(df_co2$value, xlab="Period",
     ylab="CO2 (ppm)", main="Time Series")
acf(df_co2$value, main = "ACF")
pacf(df_co2$value, main = "PACF")
hist(df_co2$value, main = "Histogram", xlab = "CO2 (ppm)")
```

*Core plot write up*
The time series plot shows an increasing average as time increases. The histogram of Co2 values also shows an non-normal distribution. Both of these point to an increasing trend in the data. The ACF plot shows persistent significant lags, even for very large lag values, this is an attribute of an AR process. Additionally, the PACF plot quickly dies, but maintain an oscillating pattern, which is an attribute of an MA process. All of these factors together point towards an ARIMA process that underlies the Co2 time series that we wish to model.

```{r model selection}
# selection with AIC
# model selection with AIC
model.aic<- df_co2 %>%
  model(ARIMA(value ~ 1 + pdq(0:10,0:2,0:10) + PDQ(1:2,1:2,1:2),
              ic="aic", stepwise=F, greedy=F))

model.aic %>%
  report()

# selection with AIcc
model.aicc<- df_co2 %>%
  model(ARIMA(value ~ 1 + pdq(0:10,0:2,0:10) + PDQ(0,0,0),
              ic="aicc", stepwise=F, greedy=F))

model.aicc %>%
  report()

# selection with BIC
model.bic<- df_co2 %>%
  model(ARIMA(value ~ 1 + pdq(0:10,0:2,0:10) + PDQ(0,0,0),
              ic="bic", stepwise=F, greedy=F))

model.bic %>%
  report()

```


*Model Interpretation*
Model selection across AIC, AICc, and BIC all chose an ARIMA model with 2 AR parameters, 4 MA parameters and linear differencing.

```{r}
sar_aic <- df_co2 %>% 
  model(ARIMA(value, ic = 'aic')) 
report (sar_aic)

sar_bic <- df_co2 %>% 
  model(ARIMA(value, ic = 'bic')) 
report (sar_bic)
```
```{r}
# testing performance of SARIMA

#plots
plot(resid(sar_bic)$.resid, main = "Residual plot")
acf(resid(sar_bic)$.resid, main = "ACF")
pacf(resid(sar_bic)$.resid, main = "PACF")
```


```{r}
# testing performance of ARIMA(2, 1, 4)

#plots
plot(resid(model.aic)$.resid, main = "Residual plot")
acf(resid(model.aic)$.resid, main = "ACF")
pacf(resid(model.aic)$.resid, main = "PACF")
```

```{r}
# forecast to the 2022
forecast_arima <- forecast(sar_bic, h = 300)
forecast_arima %>% ggplot(aes(x = index, y=.mean)) + geom_line()
```



## (3 points) Task 4a: Forecast atmospheric CO2 growth 

Generate predictions for when atmospheric CO2 is expected to be at [420 ppm](https://research.noaa.gov/article/ArtMID/587/ArticleID/2764/Coronavirus-response-barely-slows-rising-carbon-dioxide) and 500 ppm levels for the first and final times (consider prediction intervals as well as point estimates in your answer). Generate a prediction for atmospheric CO2 levels in the year 2100. How confident are you that these will be accurate predictions?

```{r}
# add prediction intervals
long_forecast_arima = forecast(sar_bic, h = 700)
long_forecast_arima[min(which(long_forecast_arima$.mean > 420)), 'index'] # first time larger than 420
```
```{r}
long_forecast_linear = forecast(linear_model, h = 700)
#long_forecast_linear[min(which(long_forecast_linear > 420)), 'index'] # first time larger than 420
co2_present_mnth
```
Dec 2041 Linear model predicts 420ppm.

# Report from the Point of View of the Present 

One of the very interesting features of Keeling and colleagues' research is that they were able to evaluate, and re-evaluate the data as new series of measurements were released. This permitted the evaluation of previous models' performance and a much more difficult question: If their models' predictions were "off" was this the result of a failure of the model, or a change in the system? 

## (1 point) Task 0b: Introduction 

In this introduction, you can assume that your reader will have **just** read your 1997 report. In this introduction, **very** briefly pose the question that you are evaluating, and describe what (if anything) has changed in the data generating process between 1997 and the present. 

## (3 points) Task 1b: Create a modern data pipeline for Mona Loa CO2 data.

The most current data is provided by the United States' National Oceanic and Atmospheric Administration, on a data page [[here](https://gml.noaa.gov/ccgg/trends/data.html)]. Gather the most recent weekly data from this page. (A group that is interested in even more data management might choose to work with the [hourly data](https://gml.noaa.gov/aftp/data/trace_gases/co2/in-situ/surface/mlo/co2_mlo_surface-insitu_1_ccgg_HourlyData.txt).) 

Create a data pipeline that starts by reading from the appropriate URL, and ends by saving an object called `co2_present` that is a suitable time series object. 

Conduct the same EDA on this data. Describe how the Keeling Curve evolved from 1997 to the present, noting where the series seems to be following similar trends to the series that you "evaluated in 1997" and where the series seems to be following different trends. This EDA can use the same, or very similar tools and views as you provided in your 1997 report. 

```{r}
# co2_present <- read.csv("co2_weekly_mlo.csv")
# co2_present <- co2_present %>%
#    group_by(year, month) %>%
#      summarise(mean = mean(average)) %>%
#      mutate(time_index = yearmonth(make_datetime(year, month))) %>%
#       as_tsibble(index = time_index)
# co2_present

# download and use the CSV

# weekly data url
weekly_url <- 'https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_weekly_mlo.txt'
# download file into weekly csv
download.file(weekly_url, "weekly.csv")
# create headings
headings = c("year", "month", "day", "decimal", "average", "ndays", "1_yr_ago", "10_yrs_ago", "increase_since_1800")
# read into and format dataframe
new_co2_w<- read.table("weekly.csv", header=FALSE, quote="\"", stringsAsFactors=TRUE)
colnames(new_co2_w) <- headings
co2_present <- new_co2_w %>%
  mutate(time_index = as.Date(make_datetime(year, month, day))) %>% as_tsibble(index = time_index)
co2_present[co2_present == -999.99] = NA
head(co2_present)
```


```{r}
co2_present %>%
  ggplot() + 
  aes(x=time_index, y=average) + 
  geom_line(color = 'steelblue') +
  labs(
    title = TeX(r'(Weekly Mean $CO_2$)'),
    subtitle = 'The "Keeling Curve"',
    x = 'Week and Year',
    y = TeX(r'($CO_2$ parts per million)')
  )
```

```{r}
# co2_present %>%
#   autoplot(mean)
summary(co2_present$average)
```

```{r}
CO2_lags <- co2_present %>% 
  gg_lag(average, geom = 'point') + 
  labs(x = "lag(value,k)", 
       title = "Lag Plots for CO2 (ppm)")


CO2_ACF <- co2_present %>% 
  ACF(average) %>% 
  autoplot() + labs(title = "The ACF for CO2 (ppm) at Muana Observatory")

CO2_lags |  CO2_ACF 
```
```{r}
CO2_PACF <- co2_present %>% 
  ACF(average, type = "partial") %>% 
  autoplot() + labs(title = "The PACF for CO2 (ppm) at Muana Observatory")
CO2_PACF
```


## (1 point) Task 2b: Compare linear model forecasts against realized CO2

Descriptively compare realized atmospheric CO2 levels to those predicted by your forecast from a linear time model in 1997 (i.e. "Task 2a"). (You do not need to run any formal tests for this task.) 




```{r}
lin_forecast <- forecast(linear_model, h=312)
lin_forecast_df <- data.frame(lin_forecast)
lin_forecast_df$index = yearmonth(row.names(lin_forecast_df))

# plot without SE
# comp_plot <- lin_forecast_df %>%
#   ggplot() + aes(x=index, y = Point.Forecast, color = "ARIMA Model") + geom_line() + 
#   geom_line(data=real_data, aes(x=time_index, y =average, color = "Actual")) +
#   ggtitle("Comparing Actual vs Predicted")
# 
# comp_plot

# plot with 80 & 95% ribbons
comp_plot <- lin_forecast_df %>%
  ggplot() + aes(x=index, y = Point.Forecast, color = "Linear Model") + geom_line() + 
  geom_ribbon(aes(ymax = Hi.80, ymin = Lo.80, fill = "80%"), alpha = 0.4, color = NA) + 
  geom_ribbon(aes(ymax = Hi.95, ymin = Lo.95, fill = "95%"), alpha = 0.2, color = NA) +
  geom_line(data=real_data, aes(x=time_index, y =average, color = "Actual")) +
  scale_fill_manual("", values = c("blue", "blue")) + 
  ggtitle("Comparing Actual vs Predicted") + ylab("co2") + xlab("Time")

comp_plot
```


```{r}

twoa_plot <- forecast_polynomial %>%
  ggplot() + aes(x=index, y = .mean) + geom_line() + ggtitle("Forecast Plot Polynomial Model")

real_data <- co2_present #%>%
  # filter(mean > 120) %>%
  # filter(year > 1997)

twob_plot <- real_data %>%
  ggplot() + aes(x=time_index, y = average) + geom_line() + ggtitle("Actual Co2 Data")

twoa_plot | twob_plot

```
It looks like the real data of Co2 levels after 1998 are significanlty higher than what was forecasted. 

## (1 point) Task 3b: Compare ARIMA models forecasts against realized CO2  

Descriptively compare realized atmospheric CO2 levels to those predicted by your forecast from the ARIMA model that you fitted in 1997 (i.e. "Task 3a"). Describe how the Keeling Curve evolved from 1997 to the present. 


```{r}

comp_plot <- forecast_arima %>%
  autoplot() + 
  geom_line(data=real_data, aes(x=time_index, y =average, color = "Actual")) +
  ggtitle("Comparing Actual vs Predicted") + ylab("co2") + xlab("Time")

comp_plot
```


```{r}

threea_plot <- forecast_arima %>%
  ggplot() + aes(x=index, y = .mean) + geom_line() + ggtitle("Forecast Plot ARIMA Model")



threea_plot | twob_plot
```
The ARIMA model performed slightly better but still underestimated the increase in co2 levels.


## (3 points) Task 4b: Evaluate the performance of 1997 linear and ARIMA models 

In 1997 you made predictions about the first time that CO2 would cross 420 ppm. How close were your models to the truth? 

In 1997 it was predicted by the ARIMA model that we would cross 420 sometime in 2031, however in reality we crossed if between 2020 and 2023.

After reflecting on your performance on this threshold-prediction task, continue to use the weekly data to generate a month-average series from 1997 to the present, and compare the overall forecasting performance of your models from Parts 2a and 3b over the entire period. (You should conduct formal tests for this task.) 

```{r}
co2_present
```


```{r}
co2_present_mnth <- co2_present %>% 
  index_by(mnth = yearmonth(as.POSIXlt(time_index, format="%Y-%m-%d"))) %>% 
  summarise(mean = mean(average))

windowed <- co2_present_mnth %>%
  filter_index("1998-1"~ "2023-12")
```

#Linear Model Eval

```{r}
lin_fore_acc <- forecast(linear_model, h = 312)
lin_fore_acc
```



```{r}
#linear model

accuracy(lin_fore_acc, windowed$mean)
```
```{r}
checkresiduals(linear_model)
```
```{r}
residuals_linear <- residuals(linear_model)
Box.test(residuals_linear, lag = 10, type = "Ljung-Box")
```


#Quadratic Model Eval

```{r}
quad_fore_acc <- forecast(quadratic_model, h = 312)
```


```{r}
accuracy(quad_fore_acc$.mean, windowed$mean)
```

```{r}
quadratic_model %>%
  augment() %>%
  ACF(.resid) %>%
  autoplot()
```
```{r}
resid.q <- quadratic_model %>%
  augment() %>%
    dplyr::select(.resid) %>%
      as.ts()

Box.test(resid.q, lag = 10, type = "Ljung-Box")
```



#Polynomial Model Eval
```{r}
poly_fore_acc <- forecast(polynomial_model, h = 312)

```



```{r}
#polynomial model

accuracy(poly_fore_acc$.mean, windowed$mean)
```
```{r}
#residuals poly
polynomial_model %>%
  augment() %>%
  ACF(.resid) %>%
  autoplot()
```



```{r}
#box test poly

resid.p <- polynomial_model %>%
  augment() %>%
    dplyr::select(.resid) %>%
      as.ts()

Box.test(resid.p, lag = 10, type = "Ljung-Box")
```

#ARIMA Model Eval

```{r}
forecast_arima_acc  <- forecast(sar_bic, h = 312)
accuracy(forecast_arima_acc$.mean, windowed$mean)
```


```{r}
#residuals arima

sar_bic %>%
  augment() %>%
  ACF(.resid) %>%
  autoplot()

```


```{r}
#box test arima 

resid.sar <- sar_bic %>%
  augment() %>%
  dplyr::select(.resid) %>%
  as.ts()

Box.test(resid.sar, lag = 10, type = "Ljung-Box")
```




## (4 points) Task 5b: Train best models on present data

Seasonally adjust the weekly NOAA data, and split both seasonally-adjusted (SA) and non-seasonally-adjusted (NSA) series into training and test sets, using the last two years of observations as the test sets. For both SA and NSA series, fit ARIMA models using all appropriate steps. Measure and discuss how your models perform in-sample and (psuedo-) out-of-sample, comparing candidate models and explaining your choice. In addition, fit a polynomial time-trend model to the seasonally-adjusted series and compare its performance to that of your ARIMA model.


```{r message = FALSE}
co2_ts <- co2_present %>%
  dplyr::select(time_index, mean) %>%
  ungroup() %>%
  dplyr::select(-year)
#co2_ts 
co2.ts <- ts(as.numeric(co2_ts$mean), start = c(1974,5), end = c(2024,3), frequency = 12)
head(co2.ts)
```


```{r}
decompose_co2 = decompose(co2.ts, "additive")
```


```{r}
adjust_co2 = decompose_co2$x - decompose_co2$seasonal
plot(adjust_co2)
```
```{r}
# adjust_co2 <- adjust_co2 %>%
#   as_tsibble()
# seasonally_adjusted_train <- adjust_co2 %>%
#   filter_index(~ "2022-3")
# 
# seasonally_adjusted_test <- adjust_co2 %>%
#   filter_index("2022-3" ~.)

#decompose_co2
```


## (3 points) Task Part 6b: How bad could it get?

With the non-seasonally adjusted data series, generate predictions for when atmospheric CO2 is expected to be at 420 ppm and 500 ppm levels for the first and final times (consider prediction intervals as well as point estimates in your answer). Generate a prediction for atmospheric CO2 levels in the year 2122. How confident are you that these will be accurate predictions?
