---
title: "W271 Group Lab"
subtitle: "Bike share demand"
author: "Annie DeForge, Hannah Abraham, Mariah Ehmke, Nora Povejsil"
output: bookdown::pdf_document2
fontsize: 11pt
geometry: margin=1in
---


# Short Questions
```{r load packages for homework 2, message=FALSE, warning=FALSE, echo=FALSE}
install.packages("corrplot")
library(corrplot)
library(tidyverse)
library(stargazer)
library(finalfit)
library(nnet)
library(car)
library(ordinal)
theme_set(theme_bw()) # set the theme (theme_set is built inside ggplot2)
```

# Political ideology (30 points)

These questions are based on Question 14 of Chapter 3 of the textbook "Analysis of Categorical Data with R" by Bilder and Loughin.

> An example from Section 4.2.5 examines data from the 1991 U.S. General Social Survey that cross-classifies people according to

> - Political ideology: Very liberal (VL), Slightly liberal (SL),Moderate (M), Slightly conservative (SC), and Very conservative (VC)
> - Political party: Democrat (D) or Republican (R)
> - Gender: Female (F) or Male (M).

> Consider political ideology to be a response variable, and political party and gender to be explanatory variables. The data are available in the file pol_ideol_data.csv.

## Recode Data (2 points)

Use the factor() function with the ideology variable to ensure that R places the levels of the ideology variable in the correct order.

```{r read in data, echo=FALSE}
pol_ideol <- read.csv("~/mids_271/spring_24_central/Labs/w271-labs/pol_ideol_data.csv",
                         header=T,na.strings=c("","NA")) 
```


```{r recode data, echo=FALSE}
pol_ideol$ideol_levels <- factor(pol_ideol$ideol, levels = c("VL", "SL", "M", "SC", "VC"))
pol_ideol$gender <- as.factor(pol_ideol$gender)
pol_ideol$party <- as.factor(pol_ideol$party)

# create new column to identify combos of party and gender
pol_ideol$segment <- ifelse(pol_ideol$gender == 'F' & pol_ideol$party == 'D', "F/D", NA)
pol_ideol$segment <- ifelse(pol_ideol$gender == 'F' & pol_ideol$party == 'R', "F/R", pol_ideol$segment)
pol_ideol$segment <- ifelse(pol_ideol$gender == 'M' & pol_ideol$party == 'D', "M/D", pol_ideol$segment)
pol_ideol$segment <- ifelse(pol_ideol$gender == 'M' & pol_ideol$party == 'R', "M/R", pol_ideol$segment)
# head(pol_ideol)
```


## Test for Independence (5 points)

Analyze the relationships between political ideology and political party and gender using basic visualizations. Afterward, generate a contingency table and assess the independence of political ideology from political party and gender.
*Comment*\
The histogram of ideology is presented first by party and then by gender. Both histograms show the largest concentration of voters in the "neutral" political ideology. Democrats are especially likely to be "neutral" as are women. Republicans are more prominent in the "slightly conservative" and "very conservative" categories. Men also are more likely to be in these categories. \
The following bar plot suggests the data were collected from an equal number of republicans and democrats, and men and women. Finally, the bar plot "Bar Plot of Ideology Categorically by Gender and Party" also suggests female democrats are more likely to fall in a liberal ideology category while male republicans will be more likely to fall in a conservative ideological category.

```{r Create a bar plot using the count variable, include = FALSE }
p <- ggplot(data=pol_ideol, aes(x=ideol_levels, y=count))+
  geom_bar(stat='identity') +
  ggtitle("Political Ideology Distribution") + 
  xlab("Ideology") + ylab("Frequency")
p
```

```{r include=FALSE}
p <- ggplot(data=pol_ideol, aes(x=party, y=count))+
  geom_bar(stat='identity') +
  ggtitle("Party Distribution") + 
  xlab("Ideology") + ylab("Frequency")
p
```

```{r include=FALSE}
p <- ggplot(data=pol_ideol, aes(x=gender, y=count))+
  geom_bar(stat='identity') +
  ggtitle("Gender Distribution") + 
  xlab("Ideology") + ylab("Frequency")
p
```

```{r gender-party divide for each ideology}
ggplot(pol_ideol, aes(fill=party, y=count, x=ideol_levels)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Party and Ideologies") +
    xlab("Ideologies")

ggplot(pol_ideol, aes(fill=gender, y=count, x=ideol_levels)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Gender and Ideologies") +
    xlab("Ideologies")
```

```{r include=FALSE}
#Plots of party and idology with gender identified in each bar plot

p1<- pol_ideol %>%
  ggplot(aes(x=ideol_levels, fill = gender)) + geom_bar() +  labs(title = "Bar Plot of Ideology Categories")
p1
p2 <- pol_ideol %>% 
  ggplot(aes(x=party, fill = gender)) + geom_bar() +  labs(title = "Bar Plot of Party Categories")
p2
```


```{r, echo=FALSE}
# barplot of relative ideology counts for each gender/party category
p1<- pol_ideol %>%
  ggplot(aes(x=segment, y= count, fill = ideol_levels)) + geom_bar(stat = "identity") +  labs(title = "Bar Plot of Ideology Categories by gender and party")
p1
```

From the visualizations we can see that of the conservative categories, there are more men. For the liberal categories there are more women. The greatest proportion of the observations are moderate, and women are more likely to be moderate.

*Comment* \
The results of the Chi-Square test for independence tests the null hypothesis $H_{0}:\pi_{ij}=\pi_{i}\pi_{j}$. The alternative hypothesis is $H_{a}:\pi_{ij}\neq\pi_{i}\pi_{j}$. We reject the null hypothesis for the test of independence between gender and ideology ($\chi^2=10.73, p<0.05$). Gender and ideology are not independent according to this test. The chi-square value is 60.905 and p<0.001 for the test of independence between ideology and party. We reject the null. The party and ideology levels are not independent. 
```{r Create a contingency table and test for the independence, echo=FALSE}
# independence of political ideology and gender
tab1 <- xtabs(count ~ ideol_levels + gender, data=pol_ideol)
tab1 # contingency table ideol and gender

test1<- chisq.test(tab1, correct=FALSE) ## chi-square test
# test1$stdres
test1

# independence of political ideology and party
tab2 <- xtabs(count ~ ideol_levels + party, data=pol_ideol)
tab2 # contingency table idol an party

test1<- chisq.test(tab2, correct=FALSE) # chi-square test
# test1$stdres
test1

```

The p-values for gender and party are both small, so the null hypothesis is that party and gender are independent from ideology is rejected, and we can conclude that ideology is dependent on these two variables.


## Regression analysis  (5 points)

Estimate a multinomial regression model and ordinal (proportional odds) regression model that both include party, gender, and their interaction. Perform Likelihood Ratio Tests (LRTs) to test the importance of each explanatory variable.

Also, test whether the proportional odds assumption in the ordinal model is satisfied. Based on this test and other results, which model do you think is more valid?

*Comment*\
We estimate the multinomial logit and ordinal regression models. The AIC for the multinomial logit model is 2491.087. It is 2484.15 for the ordinal model. 

Using Anova to test each of the explanatory variables using a $\chi^2$ test. In the multinomial logit model, only one variable, party, is significant--only as an isolated variable and not in the interaction term. In the multinomial logit model, party is positively related to a voter being in any one of the non-neutral categories. Gender does not achieve significance with a p-value of 0.06. 

In the ordinal model, both party and gender, and their interaction term is statistically significant (p<0.001). For the ordinal model, the p-value for party, gender and its interaction was small, so these variables do have an affect on ideology under this model.

We then use the proportional odds test to determine whether the odds model is appropriate for this task. While the ordinal model performed slightly better than the multinomial by AIC and number of significant coefficients, the hypothesis that the probability of the outcome increases with each level is rejected. The p-value of their significance in explaining the likelihood improvement is does not support the null. The multinomial model is our model of choice for this exercise.
```{r multinomial and ordinal regressions, echo=FALSE}
# multinomial model
mod.nomial <- multinom(ideol_levels ~ party + gender + party:gender, data=pol_ideol, weights = count)
summary(mod.nomial)
# ordinal model
mod.ord <- clm(ideol_levels ~ party + gender + party:gender, data=pol_ideol, weights = count)
summary(mod.ord)
```


Having printed the multinomial and ordinal model, we can run an ANOVA test on both.

```{r LRTs, echo=FALSE}
# anova test on multinomial model
Anova(mod.nomial, test = "LR")
# anova test on ordinal model
Anova(mod.ord, test = "Chisq")
```

Using Anova to test each of the explanatory variables in the multinomial model, the p-value is small for party, so we can conclude that is has an effect on ideology. The p-values were large for gender and the party-gender interaction effect, so for these variables we are unable to reject the null hypothesis.

For the ordinal model, the p-value for party, gender and its interaction was small, so these variables do have an affect on ideology under this model.


```{r proportional odds assumption, echo=FALSE}
# nominal test for proportional odds assumption 
nominal_test(mod.ord)
```

The p-values for all of the variables are large, we cannot reject the null hypothesis for proportional odds, so the ordinal model is not valid and we should use the multinomial model.  


## Estimated probabilities  (5 points)

Compute the estimated probabilities for each ideology level given all possible combinations of the party and gender levels.

```{r estimated probabilities for each ideology level, echo=FALSE}
# predict probability of all ideologies for female & democrat
data.f.d <- data.frame(party = 'D', gender = 'F')
pred.f.d <- predict(mod.nomial, newdata = data.f.d, type="probs", se=TRUE)

# predict probability of all ideologies for female & republican
data.f.r <- data.frame(party = 'R', gender = 'F')
pred.f.r <- predict(mod.nomial, newdata = data.f.r, type="probs", se=TRUE)

# predict probability of all ideologies for male & democrat
data.m.d <- data.frame(party = 'D', gender = 'M')
pred.m.d <- predict(mod.nomial, newdata = data.m.d, type="probs", se=TRUE)

# predict probability of all ideologies for male & republican
data.m.r <- data.frame(party = 'R', gender = 'M')
pred.m.r <- predict(mod.nomial, newdata = data.m.r, type="probs", se=TRUE)

# predicted probabilities together into table
tab_dat <- rbind(pred.f.d, pred.f.r, pred.m.d, pred.m.r)
colnames(tab_dat) <- c("VL", "SL", "M", "SC", "VC")
rownames(tab_dat) <- c("Female, Democrat", "Female, Republican", "Male, Democrat", "Male, Republican")
tab <- as.table(tab_dat)
tab
```

## Contingency table of estimated counts (5 points)

Construct a contingency table with estimated counts from the model. These estimated counts are found by taking the estimated probability for each ideology level multiplied by their corresponding number of observations for a party and gender combination.

For example, there are 264 observations for gender = "F" and party = "D". Because the multinomial regression model results in $\hat{\pi}_{VL} = 0.1667$, this model’s estimated count is $0.1667 \times 264 = 44$.

- Are the estimated counts the same as the observed? Conduct a goodness of fit test for this and explain the results.

```{r a contingency table with estimated counts, echo=FALSE}
# get observed total counts for 4 cross-categories
fd.sum <- sum(pol_ideol[which(pol_ideol$gender == "F" & pol_ideol$party == 'D'), ]$count)
fr.sum <- sum(pol_ideol[which(pol_ideol$gender == "F" & pol_ideol$party == 'R'), ]$count)
md.sum <- sum(pol_ideol[which(pol_ideol$gender == "M" & pol_ideol$party == 'D'), ]$count)
mr.sum <- sum(pol_ideol[which(pol_ideol$gender == "M" & pol_ideol$party == 'R'), ]$count)

# VL contingency
vl.prob <- unname(tab[, 1])
vl.dat <- matrix(round(c(vl.prob[1]*fd.sum, vl.prob[2]*fr.sum, vl.prob[3]*md.sum, vl.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(vl.dat) <- c("F", "M")
rownames(vl.dat) <- c("D", "R")

# SL contingency
sl.prob <- unname(tab[, 2])
sl.dat <- matrix(round(c(sl.prob[1]*fd.sum, sl.prob[2]*fr.sum, sl.prob[3]*md.sum, sl.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(sl.dat) <- c("F", "M")
rownames(sl.dat) <- c("D", "R")

# M contingency
m.prob <- unname(tab[, 3])
m.dat <- matrix(round(c(m.prob[1]*fd.sum, m.prob[2]*fr.sum, m.prob[3]*md.sum, m.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(m.dat) <- c("F", "M")
rownames(m.dat) <- c("D", "R")

# SC contingency
sc.prob <- unname(tab[, 4])
sc.dat <- matrix(round(c(sc.prob[1]*fd.sum, sc.prob[2]*fr.sum, sc.prob[3]*md.sum, sc.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(sc.dat) <- c("F", "M")
rownames(sc.dat) <- c("D", "R")

# VC contingency
vc.prob <- unname(tab[, 5])
vc.dat <- matrix(round(c(vc.prob[1]*fd.sum, vc.prob[2]*fr.sum, vc.prob[3]*md.sum, vc.prob[4]*mr.sum)), nrow = 2, ncol = 2)
colnames(vc.dat) <- c("F", "M")
rownames(vc.dat) <- c("D", "R")

# print all contingency tables
print("VL contingency table")
vl.dat
print("SL contingency table")
sl.dat
print("M contingency table")
m.dat
print("SC contingency table")
sc.dat
print("VC contingency table")
vc.dat

```

```{r include=FALSE}
# checking calculations on contingency table
# fd.sum
# fr.sum
# md.sum
# mr.sum
# 
# pred.f.d*fd.sum
# pred.m.d*md.sum
```




```{r goodness of fit test for observed vs. expected, echo=FALSE}
# GoF test VL observed with expected
vl.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'VL'), ]$count, nrow=2, ncol=2)
chisq.test(vl.dat, vl.actual)

# GoF test SL observed with expected
sl.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'SL'), ]$count, nrow=2, ncol=2)
chisq.test(sl.dat, sl.actual)

# GoF test M observed with expected
m.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'M'), ]$count, nrow=2, ncol=2)
chisq.test(m.dat, m.actual)

# GoF test SC observed with expected
sc.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'SC'), ]$count, nrow=2, ncol=2)
chisq.test(sc.dat, sc.actual)

# GoF test VC observed with expected
vc.actual <- matrix(pol_ideol[which(pol_ideol$ideol_levels == 'VC'), ]$count, nrow=2, ncol=2)
chisq.test(vc.dat, vc.actual)
```
*Comment*
The p-values are large which means that there is not enough evidence to reject the null hypothesis and conclude that the distribution of the expected and actual counts are different from each other. The estimated counts are the same as the observed counts.


## Odds ratios and confidence intervals  (8 points)

To better understand relationships between the explanatory variables and the response, compute odds ratios and their confidence intervals from the estimated models and interpret them.\

The odds ratios for a given variable, depend on the category of comparison (e.g., VL, SL, N, SC, or VL). In the multinomial logit model, the left-out gender category was female and the left-out party category was democrat. The comparison level for ideology was 'Neutral.' We compare the odds of the different coefficients for the coefficients from each VL, SL, SC, and VL level. \
Assuming the model specification 
$\log(\pi_{j}/\pi_{1}) = \beta_{j0} + \beta_{jGender}x_{Gender} + \beta_{jParty} + \beta_{jGender*Party}$ where j = SC, SL, VC, VL. \

The Odds Ratio for male versus female for the each jth level is equal to  $\frac{exp(\beta_{j0}+\beta_{j1}*(gender+c)+\beta_{j3}*party*(gender+c))}{exp(\beta_{j0}+\beta_{j1}*gender+\beta_{j3}*party*gender)} = exp(\beta_{j1}*c+\beta_{j3}*party*c)$ where c = 1 for a categorical variable. Further, gender will be equal to one as it is also a categorical variable. \

The Odds Ratio for republican versus democrat for each jth level is equal to $\frac{exp(\beta_{j0}+\beta_{j2}*(party+c)+\beta_{j3}*gender*(party+c))}{exp(\beta_{j0}+\beta_{j2}*party+\beta_{j3}*gender*party)} = exp(\beta_{j2}*c+\beta_{j3}*gender*c)$ where c = 1. 
    

`


```{r Odds ratios and confidence intervals, echo = FALSE}
#Odds Ratios 
levels(pol_ideol$ideol)
sum_mlmodel <- summary(mod.nomial)
beta_hats <- sum_mlmodel$coefficients #Coefficients without the intercept?
print("model summary")
sum_mlmodel
print("beta_hats")
beta_hats
# Neutral is comparison category 
  
#beta.hat_jr for r = 1,...6 and j = 2,3,4,5
beta.hat.sl <- beta_hats[1,2:4]
beta.hat.m <- beta_hats[2,2:4]
beta.hat.sc <- beta_hats[3,2:4]
beta.hat.vc <- beta_hats[4,2:4]

#beta_hats for individuals (GENDER), compared to democratic female
  #Slightly Conservative
sc.beta.male.rep <- exp(beta.hat.sc[1] + beta.hat.sc[2] + beta.hat.sc[3])
sc.beta.male.dem <- exp(beta.hat.sc[2])
sc.beta.fem.rep <- exp(beta.hat.sc[1])
#sc.beta.fem.dem <- 

    #Slightly Liberal
sl.beta.male.rep <- exp(beta.hat.sl[1] + beta.hat.sl[2]+ beta.hat.sc[3])
sl.beta.male.dem <- exp(beta.hat.sl[2])
sl.beta.fem.rep <- exp(beta.hat.sl[1])
#sl.beta.fem.dem <-

    #Very Conservative
vc.beta.male.rep <- exp(beta.hat.vc[1] + beta.hat.vc[2]+ beta.hat.sc[3])
vc.beta.male.dem <- exp(beta.hat.vc[2])
vc.beta.fem.rep <- exp(beta.hat.vc[1])
#vc.beta.fem.dem <-
  
  #moderate
m.beta.male.rep <- exp(beta.hat.m[1] + beta.hat.m[2]+ beta.hat.m[3])
m.beta.male.dem <- exp(beta.hat.m[2])
m.beta.fem.rep <- exp(beta.hat.m[1])
#vl.beta.fem.dem <-

# #beta_hats for individuals (PARTY)
#   #Slightly Conservative
# sc.beta.rep.male <- exp(beta.hat.sc[2] + beta.hat.sc[3])
# sc.beta.rep.fem <- exp(beta.hat.sc[2])
# sc.beta.dem.male <- exp(beta.hat.sc[3])
# #sc.beta.dem.fem <- 
#   
#   #Slightly Liberal
# sl.beta.rep.male <- exp(beta.hat.sl[2] + beta.hat.sl[3])
# sl.beta.rep.fem <- exp(beta.hat.sl[2])
# sl.beta.dem.male <- exp(beta.hat.sl[3])
# #sl.beta.dem.fem <-   
# 
#   #Very Conservative
# vc.beta.rep.male <- exp(beta.hat.vc[2] + beta.hat.vc[3])
# vc.beta.rep.fem <- exp(beta.hat.vc[2])
# vc.beta.dem.male <- exp(beta.hat.vc[3])
# #vc.beta.dem.fem <-    
# 
#   #Very Liberal
# vl.beta.rep.male <- exp(beta.hat.vl[2] + beta.hat.vl[3])
# vl.beta.rep.fem <- exp(beta.hat.vl[2])
# vl.beta.dem.male <- exp(beta.hat.vl[3])
# #vl.beta.dem.fem <- 
  


```

\
Table Odds Ratios for Gender\

Generally, there are higher odds men will be rated as either slightly or very conservative rather than moderate compared to women. Men are more likely to fall in the conservative categories instead of the moderate category if they are republicans, rather than democrats. Men were much more likely than women to be in the liberal categories instead of the neutral category conditional if they were classified as democrat. Generally, republican women than democratic women to be in the the conservative categories rather than moderate. 
```{r}
ideology_labels <- c('SL', 'SL', 'SL', 'M', 'M', 'M', 'SC', 'SC', 'SC', 'VC', 'VC', 'VC')
gender_labels <- c('Male','Male', 'Female', 'Male','Male', 'Female', 'Male','Male', 'Female', 'Male','Male', 'Female')
party_labels <- c('Rep', 'Dem', 'Rep', 'Rep', 'Dem', 'Rep', 'Rep', 'Dem', 'Rep', 'Rep', 'Dem', 'Rep' )
odds_ratio_gender <- c(sl.beta.male.rep, sl.beta.male.dem, sl.beta.fem.rep, m.beta.male.rep, m.beta.male.dem, m.beta.fem.rep, sc.beta.male.rep, sc.beta.male.dem,sc.beta.fem.rep, vc.beta.male.rep, vc.beta.male.dem, vc.beta.fem.rep)

gender_odds <- data.frame(Ideology = ideology_labels, Gender = gender_labels, Party = party_labels, OR.hat = round(odds_ratio_gender, 4))
gender_odds
```


Looking at this table, we can see that the largest odds ratios occurred for men or women Republicans who identified as slightly conservative they were 7.17 and 4.14 times, respectively, more likely to identify as slightly conservative than very liberal compared to the base comparison of a female Democrat. The smallest odds ratios were for male democrats who were less likely to identify with an ideology compared to very liberal.

```{r include=FALSE}

# odds_ratio_party <- c(sc.beta.rep.male, sc.beta.rep.fem,sc.beta.dem.male, sl.beta.rep.male, sl.beta.rep.fem, sl.beta.dem.male, vc.beta.rep.male, vc.beta.rep.fem, vc.beta.dem.male, vl.beta.rep.male, vl.beta.rep.fem, vl.beta.dem.male)
# 
# party_table_party_labels <- c('Republican', 'Republican', 'Democrat', 'Republican', 'Republican', 'Democrat', 'Republican', 'Republican', 'Democrat', 'Republican', 'Republican', 'Democrat' )
# 
# party_table_gender_labels <- c('Male', 'Female', 'Male', 'Male', 'Female', 'Male','Male', 'Female', 'Male','Male', 'Female', 'Male')
# 
# party_odds <- data.frame(Ideology = ideology_labels, Gender = party_table_party_labels, Party = party_table_gender_labels, OR.hat = round(c(sc.beta.rep.male, sc.beta.rep.fem,sc.beta.dem.male, sl.beta.rep.male, sl.beta.rep.fem, sl.beta.dem.male, vc.beta.rep.male, vc.beta.rep.fem, vc.beta.dem.male, vl.beta.rep.male, vl.beta.rep.fem, vl.beta.dem.male), 4))
# 
# party_odds

```
##Confidence Intervals for Odds Ratios 

```{r}
conf.beta <- confint(object = mod.nomial, level = 0.95)
conf.beta #Results are in 3-D array 

#Outcomes for Gender SC 
varcov <- vcov(mod.nomial)
# varcov
genderlevels1 <- c(1,1,0,1,1,0,1,1,0,1,1,0)
partylevels1 <- c(1,0,1,1,0,1,1,0,1,1,0,1)
interactlevels1 <- c(1,0,0,1,0,0,1,0,0,1,0,0)

sl.beta.male.rep.ci <- sl.beta.male.rep + qnorm(p= c(0.025, 0.975))*sqrt((varcov[2,2]+varcov[3,3]-2*varcov[2,3]))
sl.beta.male.dem.ci <- sl.beta.male.dem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[3,3]))
sl.beta.fem.rep.ci <- sl.beta.fem.rep + qnorm(p= c(0.025, 0.975))*sqrt((varcov[2,2]))
```

```{r}
#Outcomes for Gender M 
m.beta.male.rep.ci <- m.beta.male.rep + qnorm(p= c(0.025, 0.975))*sqrt((varcov[6,6]+varcov[7,7]-2*varcov[6,7]))
m.beta.male.dem.ci <- m.beta.male.dem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[7,7]))
m.beta.fem.rep.ci <- m.beta.fem.rep + qnorm(p= c(0.025, 0.975))*sqrt((varcov[6,6]))
```

```{r}
#Outcomes for Gender SC 
sc.beta.male.rep.ci <- sc.beta.male.rep + qnorm(p= c(0.025, 0.975))*sqrt((varcov[10,10]+varcov[11,11]-2*varcov[10,11]))
sc.beta.male.dem.ci <- sc.beta.male.dem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[11,11]))
sc.beta.fem.rep.ci <- sc.beta.fem.rep + qnorm(p= c(0.025, 0.975))*sqrt((varcov[10,10]))
```

```{r}
#Outcomes for Gender VC 
vc.beta.male.rep.ci <- vc.beta.male.rep + qnorm(p= c(0.025, 0.975))*sqrt((varcov[14,14]+varcov[15,15]-2*varcov[14,15]))
vc.beta.male.dem.ci <- vc.beta.male.dem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[15,15]))
vc.beta.fem.rep.ci <- vc.beta.fem.rep + qnorm(p= c(0.025, 0.975))*sqrt((varcov[14,14]))
```

```{r}
gender_odds_cis <- data.frame(Ideology = ideology_labels, Gender = gender_labels, Party = party_labels, Wald.lower = c(sl.beta.male.rep.ci[1], sl.beta.male.dem.ci[1], sl.beta.fem.rep.ci[1], m.beta.male.rep.ci[1], m.beta.male.dem.ci[1], m.beta.fem.rep.ci[1], sc.beta.male.rep.ci[1], sc.beta.male.dem.ci[1],sc.beta.fem.rep.ci[1], vc.beta.male.rep.ci[1], vc.beta.male.dem.ci[1], vc.beta.fem.rep.ci[1]), Wald.upper = c(sl.beta.male.rep.ci[2], sl.beta.male.dem.ci[2], sl.beta.fem.rep.ci[2], m.beta.male.rep.ci[2], m.beta.male.dem.ci[2], m.beta.fem.rep.ci[2], sc.beta.male.rep.ci[2], sc.beta.male.dem.ci[2],sc.beta.fem.rep.ci[2], vc.beta.male.rep.ci[2], vc.beta.male.dem.ci[2], vc.beta.fem.rep.ci[2]))

gender_odds_cis
```

For all of the OR confidence intervals, the interval is not inclusive of 0 except for female republicans who were slightly conservative, meaning that for all of the other categories, we can conclude that the difference in likelihood is significantly different than 0.

```{r include = FALSE}
# sc.beta.rep.male.ci <- sc.beta.rep.male + qnorm(p= c(0.025, 0.975))*sqrt((varcov[2,2]+varcov[3,3]-2*varcov[2,3]))
# sc.beta.rep.fem.ci <- sc.beta.rep.fem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[3,3]))
# sc.beta.dem.male.ci <- sc.beta.dem.male + qnorm(p= c(0.025, 0.975))*sqrt((varcov[2,2]))
# 
# sl.beta.rep.male.ci <- sl.beta.rep.male + qnorm(p= c(0.025, 0.975))*sqrt((varcov[6,6]+varcov[7,7]-2*varcov[6,7]))
# sl.beta.rep.fem.ci <- sl.beta.rep.fem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[6,6]))
# sl.beta.dem.male.ci <- sl.beta.dem.male + qnorm(p= c(0.025, 0.975))*sqrt((varcov[7,7]))
# 
# vc.beta.rep.male.ci <- vc.beta.rep.male + qnorm(p= c(0.025, 0.975))*sqrt((varcov[10,10]+varcov[11,11]-2*varcov[10,11]))
# vc.beta.rep.fem.ci <- vc.beta.rep.fem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[10,10]))
# vc.beta.dem.male.ci <- vc.beta.rep.fem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[11,11]))
# 
# vl.beta.rep.male.ci <- vl.beta.rep.male + qnorm(p= c(0.025, 0.975))*sqrt((varcov[14,14]+varcov[15,15]-2*varcov[14,15]))
# vl.beta.rep.fem.ci <- vl.beta.rep.fem + qnorm(p= c(0.025, 0.975))*sqrt((varcov[14,14]))
# vl.beta.dem.male.ci <- vl.beta.dem.male + qnorm(p= c(0.025, 0.975))*sqrt((varcov[15,15]))
# 

```

